---
title: "Bootstraped Clumped Isotope Calibration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bootstraped Clumped Isotope Calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8, fig.height = 6
)
```

This vignette shows you how to:

1. Calculate a bootstrapped York regression for your clumped isotope calibration dataset.
2. Apply this calibration to a set of sample replicates, with propagated calibration uncertainty.

```{r setup}
library(dplyr)
library(tibble)
library(ggplot2)
library(ggdist)
theme_set(theme_bw())
library(clumpedcalib)
set.seed(123)
```

# Bootstrapped York regression

## load calibration data

Here I show this with a very small artificial set of calibration data.
Be sure to replace with your own!

The column names must be identical to `X`, `D47`, `sd_X`, and `sd_D47`.

```{r calibration-data}
raw <- tibble::tribble(
  ~ X, ~ D47, ~ sd_X, ~ sd_D47,
  11.25,0.607,0.062,0.007,
  11.26,0.602,0.070,0.006,
  11.26,0.603,0.069,0.005,
  11.60,0.604,0.068,0.005,
  12.81,0.674,0.091,.004,
  13.28,0.672,0.092,0.012,
  13.05,0.670,0.093,0.003,
  )

# or if you have your data saved as an CSV
## raw <- readr::read_csv("dat/example_calib.csv",
##                 col_names = c("X", "D47", "sd_X", "sd_D47"))
```

## calculate bootstrapped York regression

```{r calibration-bootstrap}
# this is very much a toy example with only 100 bootstraps!
calib <- clumped_calib_boot(raw, Nsim = 100) #|>
  # for the real deal, increase Nsim to something like 1e5 and save the results
  # for re-use
  ## readr::write_csv(glue::glue("calib_clumped_boot.csv"))

# then you can read these stored results like so:
## calib <- readr::read_csv("calib_clumped_boot.csv")
```

```{r, calibrationplot}
raw |>
  ggplot(aes(x = X, y = D47)) +
  labs(x = 10^6 / T^2 ~ "(K)", y = Delta[47] ~ "(I-CDES)") +
  scale_x_continuous(sec.axis = sec_axis(
    "Temperature (°C)",
    trans = \(x) sqrt(1e6 / x) - 273.15,
    breaks = seq(0, 30, 5))) +
  geom_point() +
  geom_errorbar(aes(xmin = X - sd_X, xmax = X + sd_X)) +
  geom_errorbar(aes(ymin = D47 - sd_D47, ymax = D47 + sd_D47)) +
  # for now just draw all the lines since we're only sampling a few
  # setting alpha lower allows you to overplot the draws
  # don't forget to subset to just a few 100 though, otherwise it will be slow
  geom_abline(aes(slope = slope, intercept = intercept),
              alpha = .2, data = calib)
```

# Apply it to your data

## load sample data

I've come up with some fake replicate-level data.

```{r fakedata}
dat <- tibble::tribble(
  ~ age, ~ bins, ~ d18O_PDB_vit, ~ d13C_PDB_vit, ~ D47_final, ~ outlier, ~ identifier_1, ~ broadid,
  15.2,1,12,13,0.6,FALSE,"smp1","other",
  15.4,1,8,9,0.61,FALSE,"smp1","other",
  15.7,1,9,15,0.599,FALSE,"smp2","other",
  33.2,2,12,13,0.62,FALSE,"smp3","other",
  33.7,2,8,9,0.65,FALSE,"smp4","other",
  33.6,2,8,14,0.67,FALSE,"smp5","other",
  33.9,2,9,15,0.63,FALSE,"smp5","other",
)

# You can save your data to a CSV with the same column names and load it like so:
## dat <- readr::read_csv("fake_data.csv")

pl_raw <- dat |>
  ggplot(aes(x = age, y = D47_final, colour = factor(bins))) +
  geom_point()
pl_raw
```

## just give me the results!

This is a simple wrapper function to quickly get the end result. If you want more control, see below.

```{r bootstraped}
# calculate other "normal" summary stats if desired, like N
oth <- dat |>
  group_by(bins) |>
  summarize(n = n())

# calculate d18Osw and temp median and quantile interval
sum <- apply_calib_and_d18O_boot(data = dat,
                                 calib = calib,
                                 group = bins,
                                 Nsim = 100,
                                 output = "summary") |>
  # add N back in
  left_join(oth)
```


```{r bootplot}
# make a plot
sum |>
  ggplot(aes(x = age, y = temp)) +
  ggdist::geom_pointinterval(aes(ymin = temp.lower, ymax = temp.upper,
                                 linewidth = factor(.width))) +
  scale_linewidth_manual(values = c("0.68" = 9, "0.95" = 2), guide = "none") +
  geom_text(aes(label = paste("N =", n)), nudge_x = 1.5)
```

## more control

If you want to have more control, I recommend not using the above wrapper
function, but in stead take it one step at a time:

1. `filter_outliers()`: What it says on the tin.
2. `bootstrap_means()`: Calculate bootstrapped means for our sample bins.
3. `temp_d18Osw_calc()`: Apply the temperature calibration and use the
   clumped-derived temperature and a d18Occ--d18Osw--temperature relationship
   to calculate d18Osw.
4. `our_summary()`: Summarize by calculating the median and 68% and 95%
   confidence intervals of the mean.

You can overwrite `filter_outliers()` with your own filtering function or
comment it out if you have already cleaned up your data. It assumes there's a
logical column `outlier` and a character column `broadid` with samples equal to
`"other"`.

```{r manual}
sim <- dat |>
  filter_outliers(group = bins) |>
  bootstrap_means(group = bins, # specify which columns to use
                  age = age,
                  d13C = d13C_PDB_vit,
                  d18O = d18O_PDB_vit,
                  D47 = D47_final,
                  # obviously set Nsim way higher for real data!
                  Nsim = 100) |>
  # you can choose several d18Occ--d18Osw--temperature equations, see `equations_supported()`.
  temp_d18Osw_calc(calib = calib, equation = "KimONeil1997", Nsim = 100) |>
  our_summary(group = bins)
```

We would like to thank Alvaro Fernandez for writing and sharing the Matlab code that inspired this R package.
